---
title: "ImputationR: A Pipeline for Missing Data Imputation"
author: "Gao Jingyi"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ImputationR: A Pipeline for Missing Data Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setting, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r load_rscripts, include=FALSE}
source("../R/1.missing&correlation.R")
source("../R/2.imputation.R")
source("../R/analysis.R")
```

## Introduction

Missing data is a common challenge in data analysis that can significantly impact statistical inference, model performance, and research conclusions. The `ImputationR` package offers a comprehensive solution for missing data imputation by comparing multiple methods. The package provides a systematic workflow that includes:

1. Missing data pattern analysis
2. Correlation analysis to identify relevant variables
3. Comparison of multiple imputation methods
4. Automatic selection of the best-performing method
5. Implementation of the selected method
6. Comprehensive reporting of results

This vignette provides a detailed guide to using the `ImputationR` package, demonstrating its functionality with the `blood_storage` dataset from the `medicaldata` package.

## Installation

You can install the `ImputationR` package directly from GitHub:

```{r load_github, warning=FALSE}
# Install devtools if you haven't already
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install ImputationR from GitHub
devtools::install_github("jianvhuang/missing_data_imputation")
```

The package requires several dependencies, which will be automatically installed:

```{r load_libraries, warning=FALSE}
library(ImputationR)
# Load all dependencies with one command
ImputationR::load_imputation_dependencies()

# Create output directory
dir_output <- "../example"
cat("Output will be saved to:", dir_output, "\n")
```

## Detailed Workflow

The following sections describe the components of the `ImputationR` workflow in detail.

```{r load_data}
# Users can download the csv file from the GitHub repository and store it in their own data directory.
# This example uses the blood_storage dataset from the medicaldata package. We introduce additional missing value for demonstration purpose.
data <- read.csv("../data/blood_storage_add_missing.csv", check.names = FALSE)
```


### Data Preparation

Before imputation, it's crucial to properly prepare the data by specifying the correct variable types:

- **Continuous variables**: Numeric variables with continuous values
- **Ordinal variables**: Categorical variables with ordered levels
- **Nominal variables**: Categorical variables with unordered levels
- **Information variables** (optional): Identifier variables not to be used in imputation

The package automatically converts variables to the appropriate R data types:

```{r data_preparation, warning=FALSE}
# Define variable types
# Store information column (demographic/identifier variables)
info_vars <- NULL
info_data <- data %>% select(all_of(info_vars))

continuous_vars <- c("Median.RBC.Age", "Age", "PVol", "PreopPSA","Units", "TimeToRecurrence")
ordinal_vars <- c("RBC.Age.Group", "TVol", "T.Stage", "bGS", "sGS")
nominal_vars <- c("AA", "FamHx", "BN+", "OrganConfined", "PreopTherapy", "AnyAdjTherapy", "AdjRadTherapy", "Recurrence", "Censor")
categorical_vars <- c(ordinal_vars, nominal_vars)

# Select relevant variables
selected_vars <- c(continuous_vars, categorical_vars)
data_selected <- data %>% select(all_of(selected_vars))

# convert data types
data_selected <- data_selected %>%
  mutate(across(all_of(ordinal_vars), ~ factor(., ordered = TRUE)), 
         across(all_of(nominal_vars), as.factor)) 

lapply(data_selected[ordinal_vars], levels)

data_selected[categorical_vars] <- lapply(data_selected[categorical_vars], as.factor)
data_selected[continuous_vars] <- lapply(data_selected[continuous_vars], as.numeric)

# Check data types
str(data_selected)
```

### Missing Value Analysis

The `check_missing_values()` function analyzes the patterns of missing data in the dataset:

```{r check-missing-values, warning=FALSE}
missing_result <- check_missing_values(
  data = data_selected, 
  dir_output = dir_output
)
```

```{r show-missing-plots, echo=FALSE, fig.align='center', out.width='80%'}
missing_result$barplot
knitr::include_graphics(missing_result$patternplot)
```


This function:

1. Calculates the overall missing rate in the dataset
2. Creates a missing pattern visualization using `VIM::aggr`
3. Generates a bar plot of missing values by variable
4. Saves all visualizations to the specified output directory

Understanding the patterns of missingness is critical for selecting appropriate imputation methods. The package distinguishes between:

- **Missing Completely at Random (MCAR)**: Missing values have no relationship with any data values
- **Missing at Random (MAR)**: Missing values may be related to observed data
- **Missing Not at Random (MNAR)**: Missing values are related to unobserved data

### Correlation Analysis

The `check_correlation()` function analyzes relationships between variables:

```{r check_correlation, warning=FALSE}
cor_vars <- check_correlation(
  data = data_selected, 
  continuous_vars = continuous_vars, 
  dir_output = dir_output
)
```

```{r show-correlation-plots, echo=FALSE, fig.align='center', out.width='90%'}
cor_vars$cor_plot_cont
cor_vars$cor_plot_all
cor_vars$cor_sample_size_plot
```

This function:

Calculate the correlation coefficient using maximum sample sizes

1. Calculates correlations for continuous variables using Pearson correlation
2. Calculates correlations between all variables using appropriate methods:

   - Pearson for numeric-numeric pairs
   
   - Polychoric for categorical-categorical pairs
   
   - Polyserial for numeric-categorical pairs
   
3. Creates heatmaps for visualizing correlations
4. Identifies variables with high correlation (|r| > 0.5)

Variables with strong correlations are particularly useful for imputation, as they help predict missing values more accurately.

### Method Comparison

The `MI_func()` function (called by `run_imputation_analysis()`) compares multiple imputation methods:

```{r methods, warning=FALSE}
result <- MI_func(data_selected = data_selected, 
                  dir_output = dir_output, 
                  info_vars = NULL,
                  list_noNA = c(0.1, 0.2),
                  seed = 123,
                  niter = 2,
                  ntree = 50,
                  maxiter = 5,
                  k_values = c(3, 5, 7),
                  mice_m = 3,
                  mice_maxit = 5,
                  micer_num_trees = 50,
                  selected_features = NULL,
                  methods = c("rf", "mice")) 
```

The comparison process involves:

1. Creating complete data by removing rows with missing values
2. Simulating various missing rates (from the `list_noNA` parameter)
3. Running each imputation method on the simulated data
4. Evaluating performance using:
   - Mean Squared Error (MSE) for continuous variables
   - Accuracy for categorical variables
5. Selecting the best-performing method for the actual data

## Imputation Methods

The package implements four imputation methods:

### Random Forest (missForest)

Random Forest imputation uses the `missForest` package to build ensembles of decision trees:

- **Advantages**: Handles mixed-type data, captures non-linear relationships
- **Parameters**: 
  - `ntree`: Number of trees to grow in each forest
  - `maxiter`: Maximum number of iterations


### K-Nearest Neighbors (KNN)

KNN imputation uses the `DMwR2` package to impute values based on similar observations:

- **Advantages**: Simple, works well with locally structured data
- **Parameters**:
  - `k_values`: Number of neighbors to consider (multiple values can be tested)


### Multivariate Imputation by Chained Equations (MICE)

MICE imputation uses the `mice` package to implement an iterative approach:

- **Advantages**: Accounts for uncertainty, preserves relationships between variables
- **Parameters**:
  - `mice_m`: Number of multiple imputations
  - `mice_maxit`: Maximum number of iterations
  - `method`: Imputation method for each variable type
      - pmm (predictive mean matching) for numeric,
      - logreg (logistic regression) for binary,
      - polr (proportional odds logistic regression) for ordered,
      - polyreg (polytomous logistic regression) for unordered factors


### MICE with Random Forest (miceRanger)

miceRanger combines MICE with Random Forest using the `miceRanger` package:

- **Advantages**: Combines benefits of MICE and Random Forest
- **Parameters**:
  - `mice_m`: Number of multiple imputations
  - `mice_maxit`: Maximum number of iterations
  - `micer_num_trees`: Number of trees in each forest


## Example with Blood Storage Dataset

Let's walk through a complete example using the `blood_storage` dataset from the `medicaldata` package. This dataset contains information about red blood cell storage duration and prostate cancer outcomes.

Although the `blood_storage` dataset is an excellent example due to its diverse variable types (continuous, ordinal, and nominal), it has a very low original missing rate of only about 0.5%. 

For demonstration purposes, we'll artificially introduce additional missing values to simulate a more realistic missing data scenario with a 15% missing rate. This better represents real-world data challenges and will better demonstrate the capabilities of the imputation methods.

```{r example, warning=FALSE}
# Load required libraries
library(ImputationR)
library(dplyr)

# Run imputation analysis
result <- run_imputation_analysis(
  data = data_selected,
  continuous_vars = continuous_vars,
  ordinal_vars = ordinal_vars,
  nominal_vars = nominal_vars,
  info_vars = NULL,
  dir_output = dir_output,
  data_file_name = "blood_storage_example",
  output_file_name = "blood_imputation_report.pdf",
  # Simplified for demonstration, all these paramters can be adjusted
  list_noNA = c(0.05, 0.1, 0.15, 0.2),
  seed = 123,
  niter = 2,
  ntree = 50,
  maxiter = 5,
  k_values = c(3, 5, 7),
  mice_m = 3,
  mice_maxit = 5,
  micer_num_trees = 50,
  user_defined_features = NULL,
  methods = c("rf", "mice")
) 
```

### Understanding the Results

The `run_imputation_analysis()` function returns a list of results and generates a comprehensive report:

```{r model_results}
# Accessing results
names(result)

# Best model selected
result$best_model

# Performance metrics
result$performance
```
List the generated imputed dataset
```{r imputed_dataset}
list.files(file.path(dir_output, "imputed dataset"))
```

The final imputed datasets are saved in the `imputed dataset/` folder within the specified output directory. The number and naming of output files depend on the selected imputation method:


| Method                        | Output Files                                                                 |
|------------------------------|------------------------------------------------------------------------------|
| Random Forest (`"rf"`)       | `final_imputed_data_rf.csv`<br>→ One single imputed dataset                 |
| KNN (`"knn"`)                | `final_imputed_data_knn_n=5.csv`<br>→ One file; `n` reflects selected `K`   |
| MICE (`"mice"`)              | `final_imputed_data_mice_1.csv` to `final_imputed_data_mice_m.csv`<br>→ Multiple imputed datasets (controlled by `mice_m`) |
| miceRanger (`"mice_rf"`)     | `final_imputed_data_micerf_1.csv` to `final_imputed_data_micerf_m.csv`<br>→ Multiple imputed datasets (controlled by `mice_m`)<br>Fallback: `final_imputed_data_micerf_fallback.csv` if model fails |



The report includes:

1. **Dataset Overview**: Summary of variable types and missing values
2. **Missing Value Analysis**: Patterns of missingness and visualization
3. **Correlation Analysis**: Heatmaps and identified correlations
4. **Method Comparison**: Performance metrics for each method
5. **Imputation Results**: Comparison of data before and after imputation
6. **Implementation Details**: Parameters used

**Note:** PDF report generation requires a working LaTeX installation (e.g., via the `tinytex` package).
```{r tinytex_install}
if (!requireNamespace("tinytex", quietly = TRUE)) {
  install.packages("tinytex")
  library(tinytex)
  tinytex::install_tinytex()
} else if (!tinytex::is_tinytex()) {
  # Package is installed, but TinyTeX distribution is not
  tinytex::install_tinytex()
}
```

### Interpreting Visualizations

The package generates several visualizations to help understand the data and imputation results:

1. **Missing Pattern Plot**: Shows patterns of missing values
2. **Correlation Heatmaps**: Visualize relationships between variables
3. **Method Performance Plot**: Compares MSE and accuracy across methods
4. **Before-After Boxplots**: Compare distributions before and after imputation
5. **Before-After Density Plots**: Compare continuous variable distributions
6. **Before-After Barplots**: Compare categorical variable distributions

```{r visualization, echo=FALSE, fig.align='center', out.width='100%'}
# View the combined plot for method comparison
result$combined_plot

# View boxplots comparing before and after imputation
result$final_boxplot

# View density plots comparing before and after imputation
result$final_density_plot

# View bar plots comparing before and after imputation
result$final_barplot
```

## Advanced Usage

### Customizing Analysis Parameters

You can customize various parameters to optimize the imputation process:

```{r parameters}
# Example with customized parameters
# A list of missing data proportions to simulate (e.g., 5%, 10%, ..., 50%)
list_noNA = c(0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.45, 0.5)

# Random seed to ensure reproducibility
seed = 123

# Number of iterations to repeat the entire simulation process
niter = 10

# number of trees to grow in each forest.
ntree = 100

# maximum number of iterations (in random forest) to be performed given the stopping criterion is not met beforehand.
maxiter = 10

# A vector of K values to test for KNN imputation (e.g., K = 3, 5, 7)
k_values = c(3, 5, 7)

# Number of multiple imputations to perform in MICE
mice_m = 5

# A scalar giving the number of iterations
mice_maxit = 10

# The number of decision trees in miceForest
micer_num_trees = 100

# Methods used to simulate imputation
methods = c("rf", "knn", "mice", "mice_rf")
```

## Conclusion

The `ImputationR` package provides a comprehensive framework for missing data imputation. By comparing multiple methods and providing detailed visualizations, it helps researchers select the most appropriate imputation approach for their specific dataset.

Key benefits of the package include:

1. **Comprehensive workflow**: From data preparation to final report
2. **Method comparison**: Objective selection of the best-performing method
3. **Detailed visualizations**: For understanding missing patterns and imputation results
4. **Customizability**: Multiple parameters for optimization


